---
title: R client for the calcite demonstrator
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
package: zircon
date: "Revised: October 7, 2022"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Using the calcite R client}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(BiocStyle)
self <- Githubpkg("ArtifactDB/calcite-R", "calcite");
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

`r self` implements a simple R client for interacting with the **calcite** demonstrator API.
**calcite** provides an ExperimentHub-like data store for Bioconductor objects based on the [schemas here](https://github.com/ArtifactDB/calcite-schemas).
Its purpose is to demonstrate how the ArtifactDB framework can be easily adapted for other applications.

# Fetching objects from the backend

The `fetchObject()` function will load an R object from the **calcite** backend, given the object's identifier:

```{r}
library(calcite)
(id <- exampleID())
obj <- fetchObject(id)
obj
```

We can extract the ExperimentHub-like metadata using `objectAnnotation()`:

```{r}
str(objectAnnotation(obj))
```

More complex objects can be loaded if the corresponding [**alabaster**](https://github.com/ArtifactDB/alabaster.base) packages are installed.
For example, we can load [`SingleCellExperiment`](https://bioconductor.org/packages/SingleCellExperiment) objects if `r Githubpkg("ArtifactDB/alabaster.sce")` is installed.

```r
fetchObject("test:my_first_sce@v1")
```

```
## class: SingleCellExperiment
## dim: 20006 3005
## metadata(1): .internal
## assays(1): counts
## rownames(20006): Tspan12 Tshz1 ... mt-Rnr1 mt-Nd4l
## rowData names(1): featureType
## colnames(3005): 1772071015_C02 1772071017_G12 ... 1772066098_A12
##   1772058148_F03
## colData names(10): tissue group # ... level1class level2class
## reducedDimNames(0):
## mainExpName: endogenous
## altExpNames(2): ERCC repeat
```

# Saving objects to the backend

Given some Bioconductor objects, we can annotate them with ExperimentHub-like metadata:

```{r}
library(S4Vectors)
df1 <- DataFrame(A=runif(10), B=rnorm(10), row.names=LETTERS[1:10])
df1 <- annotateObject(df1, 
    title="FOO",
    description="Ich bien ein data frame",
    maintainers="Aaron Lun <infinite.monkeys.with.keyboards@gmail.com>",
    species=9606,
    genome=list(list(id="hg38", source="UCSC")),
    origin=list(list(source="PubMed", id="123456789"))
)
```

Then we save them into a "staging directory":

```{r}
staging <- tempfile()
dir.create(staging)
saveObject(df1, staging, "df001")
list.files(staging, recursive=TRUE)
```

Any name can be used for the objects, and multiple objects can be saved into the same directory.
Objects can even be saved into subdirectories:

```{r}
df2 <- DataFrame(A=runif(10), B=rnorm(10), row.names=LETTERS[1:10])
df2 <- annotateObject(df1, 
    title="BAR",
    description="Je suis une data frame",
    maintainers=list(list(name="Darth Vader", email="vader@empire.gov")),
    species=10090,
    genome=list(list(id="GRCm38", source="Ensembl")),
    origin=list(list(source="GEO", id="GSE123456"))
)

dir.create(file.path(staging, "variants"))
saveObject(df2, staging, "variants/df002")
list.files(staging, recursive=TRUE)
```

Once we're done with staging, we're ready to upload.
We pick a project name and version and call the `uploadDirectory()` function.
This will prompt us for a [GitHub personal access token](https://github.com/settings/tokens) to authenticate into the backend, if we haven't supplied one already.

🚨🚨🚨 **ALERT!**
For cost and security reasons, the demonstrator instance only allows a certain set of GitHub users to perform new uploads.
If you're not on the [list](https://github.com/ArtifactDB/calcite-worker/master/tree/wrangler.toml), you'll have to ask someone to create a stub project for you before you can use `uploadDirectory()`.
🚨🚨🚨

```{r, eval=FALSE}
# Setting an expiry date of 1 day in the future, to avoid having lots of
# testing projects lying around in the data store.
uploadDirectory(staging, 
    project="test-vignette", 
    version=as.character(Sys.Date()), # using the date as the version.
    expires=1)
```

# `DelayedArray` wrappers for calcite

For arrays, `r self` offers some special behavior during loading:

```{r}
mat <- fetchObject("test:my_first_sce/assay-1/matrix.h5@v1")
mat
```

The `CalciteArray` object is a [`DelayedArray`](https://bioconductor.org/packages/DelayedArray) subclass that remembers its ArtifactDB identifier.
This provides some optimization opportunities during the save/upload of the unmodified array.
For example, let's wrap `mat` inside an annotated list and save it:

```{r}
obj <- List(thingy=mat)
obj <- annotateObject(obj,
    title="FOO",
    description="I am a list",
    maintainers="Aaron Lun <infinite.monkeys.with.keyboards@gmail.com>",
    species=9606,
    genome=list(list(id="hg38", source="UCSC")),
    origin=list(list(source="PubMed", id="123456789"))
)

dir <- tempfile()
dir.create(dir)
saveObject(obj, dir, "bar")
```

If we were inclined to dig around in the `bar` directory, we would see that the `array` data in `child-1` doesn't actually contain any data, but just a link to the existing ArtifactDB identifier.
This instructs `uploadDirectory()` to also skip the upload and create a link on the backend.

```{r}
zircon::extractLinkedID(dir, "bar/child-1/array")
```

In this manner, we can save compute time, upload bandwidth and storage space when creating new versions of an object.
The biggest beneficiary of this functionality is the `SummarizedExperiment`, where we can update metadata, row/column annotations, etc. at low cost provided that the underlying assay data does not change.

By default, the `CalciteArray` is created with file-backed arrays from the `r Biocpkg("HDF5Array")` package.
This ensures that it can be easily manipulated in an R session with a small memory footprint.
However, for analysis steps that actually use the array contents, it is best to convert the `CalciteArray` into an in-memory representation to avoid repeated disk queries:

```{r}
smat <- as(mat, "dgCMatrix")
str(smat)
```

# Searching for objects

🚧🚧🚧 **Coming soon** 🚧🚧🚧

# Advanced usage

The **calcite** API is just another ArtifactDB instance, so all methods in the `r Githubpkg("ArtifactDB/zircon-R")` package can be used.
For example, we can directly fetch the metadata for individual components:

```{r}
library(zircon)
meta <- getFileMetadata(exampleID(), url=restURL())
str(meta$data_frame)
```

We can inspect the permissions for a project:

```{r}
getPermissions("test", url=restURL())
```

And we can pull down all metadata for a particular version of a project:

```{r}
v1.meta <- getProjectMetadata("test", version="v1", url=restURL())
length(v1.meta)
```

# Session information {-}

```{r}
sessionInfo()
```
